An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]

### 2025-01-26 paper/essay.qmd
Converted the paper into an essay for submission to Critical AI

---
title: "An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: What Role for the Human in Higher Education?"
format: docx
bibliography:
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

## Abstract

The use of Generative Artificial Intelligence (AI) in higher education has been a sensational topic since the launch of ChatGPT and similar tools. AI has been presented as an everything tool that can do a fabulous array of tasks, from creating content to grading to providing feedback to students. This presentation of AI is misleading, and there is a need to examine the claims about what the technology can do and what it actually does in the context of teaching and learning. In this essay I scan the academic and gray literature, along with media reports regarding AI, to show how discussions about AI hinge on making a large number of unverified claims (Gish Gallop) while presenting the technology as inevitable (a False Dilemma). However, once we examine what AI does, and particularly what it does for whom and to whom, a picture emerges of a technology that is intrinsically inimical to the principles that most institutions of higher education claim to hold dear, including the best interest of their students.

@@

AI technologies in education "offer new tools and applications that have the potential to transform traditional teaching and learning methods" [@adiguzel_revolutionizing_2023, p. 1], and have "the potential to fundamentally change the classroom experience and the knowledge and skills outcomes of our students" [@dwivedi_opinion_2023, p.20]. Furthermore, the technology "is here to stay...[t]herefore, teachers and students must develop the specific digital competencies needed to use such tools in ways that are pedagogically beneficial and ethical" [@kohnke_chatgpt_2023, p. 546].

AI is painted as revolutionary: an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, while leading to _efficiencies_^["Efficiencies" being an euphemism for cutting costs [@oconnell_let_2012, p. 64], usually by finding "redundancies" rather than better things for people to do.]. Higher education institutions were compelled to get in on the act, forming high-level advisory committees and drafting guidelines for university communities to leverage AI [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].

But hammers are powerful tools, and so are thousands of other instruments found at universities. And yet not all of them get high-level university committees to advice on their use. Just what is it about AIs that requires pan-university consultations on how to use them? To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, also examine what it actually does.

@@

A regression model uses as inputs a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable) to find a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables). The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model).

@fig-bivariate-model is an example of the simplest regression model with only one independent variable. The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$. This model needs only two parameters: an intercept (the conditional mean $\hat{y}$ when $x = 0$), and the slope of the line (the rate of change of $\hat{y}$ with respect to $x$). Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.

```{r simulate-simple-example}
set.seed(13253)

b0 <- 1.0
b1 <- 1.5

df <- data.frame(x = b0 + b1 * runif(100, 
                                0, 
                                10),
                 e = rnorm(100),
                 type = "training data") |>
  mutate(y = x + e)

x <- df |> pull(x)
y <- df |> pull(y)

mod <- lm(y ~ x, 
          data = df)
```

```{r plot-simulated-data, message=FALSE, out.width="70%", fig.cap="Bivariate Regression Model"}
#| label: fig-bivariate-model

ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="70%", fig.cap="Training data and model output compared"}
#| label: fig-data-vs-preds-01

df_pred <- data.frame(x = x)

y_hat <- predict(mod, df_pred)

df_pred_lr <- df_pred |> 
  mutate(y = y_hat,
         type = "regression output")

df_pred_all <- rbind(df |>
                   select(x, y, type),
                 df_pred_lr |>
                   select(x, y, type))

ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Neural networks do something similar but using a net-like architecture. For instance, @fig-nn1 shows a relatively simple neural network trained using using the same data as above. This network is not "deep", as it has only one "hidden layer" (the two neurons between the input and the output layers). In comparison to the linear regression and its two parameters, this neural network uses seven parameters that need to be pre-loaded with random values, which makes the results of the model contingent on the starting conditions of the training process.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

```{r train-shallow-neural-network}
# Train a neural network using only two neurons in one hidden layer. The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node. The default activation function is the logistic function.
set.seed(33225433)

nn1 <- neuralnet(f, 
                data = df, 
                hidden = c(2), 
                linear.output = T) 
```

```{r save-shallow-neural-network}
# For some reason, the figure cannot be saved directly, so I saved it manually using screen capture
plot(nn1)
```

```{r plot-shallow-neural-network, out.width="70%", fig.cap="Simple Example of a Shallow Neural Network"}
#| label: fig-nn1

knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
```

```{r shallow-neural-network-predictions}
y_hat <- predict(nn1,
                 df_pred)

df_pred_nn1 <- df_pred |> 
  mutate(y = y_hat,
         type = "nn1 output")

df_pred_all <- rbind(df_pred_all |>
                   select(x, y, type),
                 df_pred_nn1 |>
                   select(x, y, type))
```

Unlike a linear regression model, the predictions of the neural network can be non-linear (@fig-data-vs-preds-02 ). This flexibility is achieved by being a less parsimonious model (it uses more parameters).

```{r data-vs-predictions-comparison-2, out.width="70%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
#| label: fig-data-vs-preds-02

ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

More complex neural networks can be designed. @fig-nn2 shows a deeper network, with two hidden layers of five neurons each. This model uses forty-six parameters. 
```{r train-deeper-neural-network}
# Train a neural network using only two neurons in one hidden layer. The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node. The default activation function is the logistic function.
set.seed(242)
nn2 <- neuralnet(f, 
                data = df, 
                hidden = c(5, 5), 
                linear.output = T)
```

```{r save-plot-deeper-neural-network}
# For some reason, the figure cannot be saved directly, so I saved it manually using screen capture
plot(nn2)
```

```{r plot-deeper-neural-network, out.width="70%", fig.cap="Example of a Deeper Neural Network"}
#| label: fig-nn2

knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
```

```{r}
y_hat <- predict(nn2,
                 df_pred)

df_pred_nn2 <- df_pred |> 
  mutate(y = y_hat,
         type = "nn2 output")

df_pred_all <- rbind(df_pred_all |>
                   select(x, y, type),
                 df_pred_nn2 |>
                   select(x, y, type))

```

The increased complexity leads to greater flexibility, and we can see that the model's predictions resemble more closely the actual data (@fig-data-vs-preds-03).

```{r data-vs-predictions-comparison-3, out.width="70%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
#| label: fig-data-vs-preds-03

ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
# How do these models perform?
mad_lr <- data.frame(df |>
             mutate(y_hat = df_pred_lr$y)) |>
  summarize(mse = sum(abs(y - y_hat)/n()))

mad_nn1 <- data.frame(df |>
             mutate(y_hat = df_pred_nn1$y)) |>
  summarize(mad = sum(abs(y - y_hat)/n()))

mad_nn2 <- data.frame(df |>
             mutate(y_hat = df_pred_nn2$y)) |>
  summarize(mad = sum(abs(y - y_hat)/n()))

delta_mad_nn1 <- (mad_nn1 - mad_lr)/mad_lr * 100

delta_mad_nn2 <- (mad_nn2 - mad_lr)/mad_lr * 100
```

Indeed, the first neural network is "better" than the linear regression in that it reduces the typical error by `r paste0(prettyNum(abs(delta_mad_nn1), digits = 2), "%")`, whereas the deeper network reduces the typical error by `r paste0(prettyNum(abs(delta_mad_nn2), digits = 2), "%")`, also with respect to the linear regression^[The "typical error" here is the mean absolute deviation of the predictions.].

There is a limit as to how flexible a regression model can be. Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model. Roughly, every additional parameter needs one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and predicts each value of the independent variable in the training dataset perfectly. The model will fail to train if there are more parameters than data points.

@@

This is an utterly unrealistic expectation. Ideally, students should have the opportunity to discuss their drafted work with instructors before final submission (or could ask members of their study groups to proof-read their work, among many other ways to get feedback before submission). A conscientious instructor will invest time reading the drafted work to provide quality feedback. There is, however, a more insidious aspect about getting feedback from an AI...who, actually, provides the feedback in this case? Whose data are used to train the machine that provides the feedback? Who fine-tunes the machine that provides the feedback? Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided? A poignant example of this dilemma involves AI-based face recognition and online proctoring, as related by @hill_accused_2022.

@@

Leaving the technology fetish about classroom applications aside, Mollick and Mollick are well aware of the ethical concerns around using AI (see p. 6), including that "[m]odels are trained in ways that may violate copyright", that exploit the "efforts of low-wage workers in precarious conditions for part of their training", and the fact that "[m]odels are trained on biased data and can produce either subtly or overtly biased results". And yet, they choose to ignore every single one of these concerns for the sake of innovation. Machine-washed plagiarism that exploits low-wage workers while using inordinate amounts of energy is innovative only in a way that offends "integrity" [@office_of_the_president_mcmaster_university_mission_2024], "individual human rights...equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992], and "sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].

But what if the datasets and training processes used to develop GenAIs were unimpeachable, would the practice of generating learning materials using AIs then be "innovative"? Alas, no. As should be clear by now, any teaching materials generated in this way can only aspire to be average given the prompts used to get the machine going, and no one should buy the delusion that the use of this technology makes them innovators. In any case, the list of remedial measures required to verify that the materials generated are not nonsense is so long, that it is hard to see what is being gained [see @mollick_instructors_2024, pp. 4-5].

@@

Computers have been doing mathematics reliably since at least the [Antikythera mechanism](https://en.wikipedia.org/wiki/Antikythera_mechanism) (ca. 178 BC),. Electronic computers have been doing mathematics for decades now, and neural networks have been used for decades to do data analysis. There is nothing new about that.  But the promise rather is that we will be able to do data analysis without having to _learn_ how to do data analysis. The idea that a chatbot could be used to do data analysis betrays a profound misunderstanding of both what chatbots do and what data analysis entails. Ask a chatbot to do some arithmetic: it will produce something that is right on average. But average is not always right. And seeing how AIs are now being trained on the dregs of the internet, it should be no surprise when an AI Overview responds to the prompt "2 + 2" with "potato".

```{r flickr-potato, out.width="70%", fig.cap="Data analysis using a chatbot? Maybe not such a good idea"}
knitr::include_graphics(glue::glue(here::here(), "/images/potato.png"))
```

@@

From a different perspective, even the average output of an AI should be treated with caution. As noted [above](#really-what-do-they-do), the results of a neural network depend on the weights selected at random at the start of a training process. Change those weights, the results will change. As well, even small variations in the prompt can be impactful in models with billions or trillions of parameters. In statistical terms, any output is but one _realization_ of a random process. For the results to be useful, we should calculate the average of a large number of responses. Given the cost of training and querying LLMs, this is prohibitively expensive.

@@

It is important to note that, even if prompt "engineering" offered some value, the skill is tied to a singular technology that offers little transferability potential. @mollick_instructors_2024 demonstrate how prompt "engineering" is not even transferable _between_ different LLMs (see p. 5 and p. 36). Good skills transfer between contexts. Learning to do effective bibliographical queries transfers between libraries. Learning to identify relevant content works for reading a novel, a book of history, or a text of econometrics. Writing, the slow and demanding process of converting thoughts into words, is essential to create a style of communication that is unique to each person. It might be argued, in fact, that writing _is_ learning [@klein_trends_2016].

An AI is like a crutch, but what we are being promised is an All-Terrain Vehicle, however one that the user will never own, and cannot hope to fully master due to its blackbox nature (neural networks are blackboxes and their development is also a blackbox wherein things can more easily change based on the interests of their corporate owners than in response to the interests of their relatively small customers^[And this refers not only to individuals. As an example, in 2022 McMaster University, a major Canadian university, operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD). Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022. To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue.].

<!--
McMaster: 842,899,000 CAD
Microsoft: 198,000,000,000 USD == 257,776,200,000 CAD

https://www.exchange-rates.org/exchange-rate-history/usd-cad-2022
Avg. exchange rate in 2022: 1.3019 CAD/US

This is how much bigger Microsoft is compared to McMaster. 
(198000000000 * 1.3019)/842899000

Put another way: If McMaster gave ALL OF ITS REVENUE to Microsoft, it would only account for 842899000 / (198000000000 * 1.3019) or less than one third of a percentage point of Microsoft's revenue.
--->

@@

Improvement in the performance of AIs has relied on using bigger datasets. This has led to a rush, as developers train their models using bigger datasets^[A rush that feels as frenetic as any gold rush in the past. However, while the gold rushes of the 19th century were relatively democratic due to their low entry costs, the current data rush has enormous entry costs and is dominated by large corporations and venture capital.]. But, despite the large number of text items and images available in digital form, the supply is finite.

As of this writing, the few big players that can afford to grab the data, are mostly done with it, using morally dubious and possibly illegal means [@clark_ex-amazon_2024; @milmo_zuckerberg_2025; @smalley_linkedin_2025]. And so, as the fountain of human-generated content has been reduced to a drip, there has been a turn towards what might be the last untapped drops of information left on the internet.

OpenAI, a global leader in data grabbing, resorted to data from Stack Overflow [@stack_overflow_stack_2024], a website that [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges". Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the deal, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not_ and _cannot_ do. It is an open question how long Stack Overflow will remain a community-based space after it started treating its members as unwitting, and even unwilling providers of data for OpenAI [@grimm_stack_2024].

Google, for its part, appealed to Reddit for training data to the tune of \$60 million per year [@roth_google_2024]. As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures that include a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024].

```{r google-reddit, out.width="70%", fig.cap="Google's AI Overview result was traced back to a post by redditor Fucksmith"}
knitr::include_graphics(glue::glue(here::here(), "/images/google-cheese-glue.png"))
```

Other AI Overview results included a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article published by the satirical periodical [The Onion](https://www.theonion.com/geologists-recommend-eating-at-least-one-small-rock-per-1846655112).

```{r google-the-onion, out.width="70%", fig.cap="Google's AI Overview result was traced back to an article in The Onion"}
knitr::include_graphics(glue::glue(here::here(), "/images/google-the-rock.jpg"))
```

These, and numerous other hapless results, are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024]. To what extent can more money improve GenAIs? Probably not by much, since a key constraint is the amount of data available for model training [@villalobos_position_2024]. And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data. As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the world's most thoroughly vetted form of writing [@cabanac_tortured_2021; @maiberg_ai_2024].

What can we expect as future GenAI models begin to be trained using data that was generated by previous GenAIs? Revisiting the examples in a preceding section ([But really, what do AIs do?](#really-what-do-they-do)) we can see that the predictions of a model are more uniform than the original data used for training: the outputs of a model will be less varied, less diverse, and more norm-conforming. The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs. We may be witnessing _Peak ChatGPT_, and future model versions, instead of improving, may begin to regress to duller versions of their own previous outputs. To compound things, as the backlash against an exploitative data rush is suuported by new tools to generate data that are deleterious to the training process of a GenAIs [@salman_raising_2023], including _PhotoGuard_ [@heikkila_this_2023], _Nightshade_ [@shan_nightshade_2024], and _Shadowcast_ [@xu_shadowcast_2024].

All in all, the evidence suggests that the scarcity of fresh data, paired with more abundant model-generated and/or poisoned data, will make the performance of these machines stagnate or deteriorate. Which begs the following questions. How useful will AIs be in the future? How persistent will skills like prompt "engineering" be? Will learning how to prompt an AI be worth sacrificing the opportunity to develop transferable skills like coding, writing, and thinking? Who, after all, stands to benefit the most by the form of deskilling that AIs encourage?

## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-for-corporations}

AIs do not seem to do much of value for students and teachers, as they are wasteful at best and harmful at worst. But the technology must do something for someone, otherwise it would not be ubiquitous. So what do AIs do for the the organizations that develop them? So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities^[Some observers believe that the AI fever has become detrimental to actual innovation [@siegel_ai_2023; @vinsel_dont_2023].]. Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.] [@tricot_venture_2021].

Alas, the industry has not necessarily made money for investors yet.

@berber_peter_2024, for instance, reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue. @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google. At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 monthly queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer. And a large number of users do not pay: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023]. The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing  to the Global South the most disturbing, degrading, and damaging aspects of training AIs [@nix_analysis_2024].

@@

## What do GenAIs do to everyone and everything else? {#what-do-ais-do-to-everyone-else}

### AI is an eminently extractive and exploitative industry

ChatGPT 4.o was announced on May 13, 2024, with great fanfare and a flashy display of the model's new capabilities, including functions to respond vocally to prompts, and to identify images via a mobile device's camera. 4.o's voice, in particular, caused considerable titillation [@roose_is_2024], modeled as it was after the digital personal assistant of the movie _Her_ (Sam Altman's favorite movie by his own account). The background story of 4.o's voice is more sordid: Altman reportedly tried to hire Scarlett Johannson, who played the AI assistant in the movie, to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when she declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways. This voice was retired after Ms. Johansson threatened legal action [@allyn_scarlett_2024].

Alas, very few have the wealth and social standing of Scarlett Johansson to challenge OpenAI. Most people in the world lack that power and the AI industry has not been shy to exploit the imbalance [@roberts_commercial_2016; @germain_google_2023; @weatherbed_adobes_2024].

Besides being extractive, the AI industry is exploitative in darker ways. ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified. Trained on horrible content, it was hard to sell a ChatGPT prone to racist, sexist, obscene outbursts. Fortunately for OpenAI, companies like facebook had already hit on a solution of sorts to reduce the cost of moderation. The idea was basically to train AIs to detect content that was unpalatable. Indeed, they had found that feeding an AI with labeled examples of all the worst that exists on the internet (e.g., violence, hate speech, child sexual abuse) could produce tools that learned to detect various manifestations of such toxicity, to filter it out before it reached users. To obtain the essential labeled inputs, OpenAI sent tens of thousands of samples to Kenya, with text that described murder, suicide, bestiality, self-harm, incest, torture, and child sexual abuse situations in graphic detail. A sanitized ChatGPT is only possible through the labor (with wages around $1.32 and $2 per hour) of workers who were exposed to the worst of the internet. As one worker tasked with reading and labeling text for OpenAI told TIME, "he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child. 'That was torture,' he said. 'You will read a number of statements like that all through the week. By the time it gets to Friday, you are disturbed from thinking through that picture'" [@perrigo_2_2023].

### Privacy issues

>>>>

---
title: "An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]"
format: docx
bibliography:
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

<!--
ABSTRACT
The use of Generative Artificial Intelligence (AI) in higher education has become a sensational topic. AI has been presented as an everything tool that can do a fabulous array of tasks, from creating content to grading to providing feedback to students. This presentation of AI is misleading, and there is a need to examine the claims about what the technology can do and what it actually does in the context of teaching and learning. In this essay I scan the academic and gray literature, along with media reports regarding AI, to show how discussions about AI hinge on making a large number of unverified claims (Gish Gallop) while presenting the technology as inevitable (a False Dilemma). However, once we examine what AI does, and particularly what it does for whom and to whom, a picture emerges of a technology that is intrinsically inimical to the principles that most institutions of higher education claim to hold dear, including the best interest of their students.
-->

@@

AI is painted as revolutionary in education: an _innovative_ and _exciting_ technology, that moreover can lead to juicy _efficiencies_^["Efficiencies" being an euphemism for cutting costs [@oconnell_let_2012, p. 64], usually by finding "redundancies" rather than better things for people to do.]. Higher education institutions were compelled to get in on the act, forming high-level advisory committees and drafting guidelines for university communities to leverage AI [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].

Among the thousands of other tools used in universities, just what is it about AIs that requires pan-university consultations on how to use them? To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, also examine what it actually does.

@@

A regression model uses as inputs a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable) to find a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables). To regress the inputs to a conditional mean some criterion needs to be adopted, for example, that the conditional mean minimizes the error of the model).

Neural networks do something similar but using a net-like architecture that allows the model to make non-linear predictions. This flexibility is achieved by using more parameters, but has limits: while each additional parameter makes the model more flexible, it requires one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and predicts each value of the independent variable in the training dataset perfectly.

@@

A conscientious instructor who is not overextended will invest time reading drafted work to provide quality feedback. There is, however, a more insidious aspect about getting feedback from an AI...who, actually, provides the feedback in this case? Whose data are used to train the machine that provides the feedback? Who fine-tunes the machine that provides the feedback? Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided? A poignant example of this dilemma involves AI-based face recognition and online proctoring, as related by @hill_accused_2022.

@@

Leaving the technology fetish about classroom applications aside, Mollick and Mollick are well aware of the ethical concerns around using AI (see p. 6), including copyright violations, exploitation of low-wage workers, as well as the use of biased data. And yet, they choose to ignore every single one of these concerns for the sake of innovation. Exploitative, world-burning machine-washed plagiarism is innovative only in a way that offends "integrity" [@office_of_the_president_mcmaster_university_mission_2024], "individual human rights...equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992], and "sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].

But what if the datasets and training processes used to develop GenAIs were unimpeachable, would the practice of generating learning materials using AIs then be "innovative"? Alas, no. As should be clear by now, AI-generated teaching materials can only aspire to be average given the prompts used, and no one should buy the delusion that the use of this technology makes them innovators. In any case, the list of remedial measures required to verify that the materials generated are not nonsense is so long, that it is hard to see what is being gained [see @mollick_instructors_2024, pp. 4-5].

@@

Computers have been doing mathematics reliably since at least the [Antikythera mechanism](https://en.wikipedia.org/wiki/Antikythera_mechanism) (ca. 178 BC),. Electronic computers have been doing mathematics for decades now, and neural networks have been used for decades to do data analysis. There is nothing new about that.  But the promise rather is that we will be able to do data analysis without having to _learn_ how to do data analysis. The idea that a chatbot could be used to do data analysis betrays a profound misunderstanding of both what chatbots do and what data analysis entails. Ask a chatbot to do some arithmetic: it will produce something that is right on average. But average is not always right. And seeing how AIs are now being trained on the dregs of the internet, it should be no surprise when an AI Overview responds to the prompt "2 + 2" with "potato" (@fig-flickr-potate).

```{r flickr-potato, out.width="70%", fig.cap="Data analysis using a chatbot? Maybe not such a good idea"}
#| label: fig-flickr-potato

knitr::include_graphics(glue::glue(here::here(), "/images/potato.png"))
```

@@

@@

It is important to note that, even if prompt "engineering" offered some value, it is a skill that offers little transferability potential. @mollick_instructors_2024 demonstrate how prompt "engineering" is not transferable _between_ different LLMs (see p. 5 and p. 36). Learning to do effective bibliographical queries transfers between libraries. Learning to identify relevant content works for reading a novel, a book of history, or a text of econometrics. Writing, the slow and demanding process of converting thoughts into words, is essential to create a style of communication that is unique to each person. It might be argued, in fact, that writing _is_ learning [@klein_trends_2016].

AIs are presented as all-terrain vehicles, but are more like a crutch, however one that the user will never own, and cannot hope to fully master due to its black-box nature [@judelson_restaging_2024]

@@

Improvement in the performance of AIs has relied on using bigger datasets. This has led to a rush, as developers train their models using bigger datasets^[A rush that feels as frenetic as any gold rush in the past. However, while the gold rushes of the 19th century were relatively democratic due to their low entry costs, the current data rush has enormous entry costs and is dominated by large corporations and venture capital.]. But, despite the large number of text items and images available in digital form, the supply is finite and, like an overexploited river^[As of this writing, the few big players that can afford to grab the data, are mostly done with it, using morally dubious and possibly illegal means [@clark_ex-amazon_2024; @milmo_zuckerberg_2025; @smalley_linkedin_2025].], human-generated content has been reduced to a drip [q.v., @delacroix_sustainable_2024]. In response to this, there has been a turn towards what might be the last untapped drops of genuine human-generated information left on the internet, often to the disgust of humans there^[OpenAI resorted to data from Stack Overflow [@stack_overflow_stack_2024], a website that [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges". Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the deal, and has come to believe that the terms were abusive].

Google, for example, appealed to Reddit for training data to the tune of \$60 million per year [@roth_google_2024]. As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures that include a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024].

```{r google-reddit, out.width="70%", fig.cap="Google's AI Overview result was traced back to a post by redditor Fucksmith"}
#| label: fig-google-reddit

knitr::include_graphics(glue::glue(here::here(), "/images/google-cheese-glue.png"))
```

Other AI Overview results included a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article published by the satirical periodical [The Onion](https://www.theonion.com/geologists-recommend-eating-at-least-one-small-rock-per-1846655112) (@fig-google-the-onion).

```{r google-the-onion, out.width="70%", fig.cap="Google's AI Overview result was traced back to an article in The Onion"}
#| label: fig-google-the-onion

knitr::include_graphics(glue::glue(here::here(), "/images/google-the-rock.jpg"))
```

These results are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024]. To what extent can more money improve GenAIs? Probably not by much, since a key constraint is the amount of data available for model training [@villalobos_position_2024]. And dismal as this is, the outlook is even worse: as GenAIs become more widely used, they produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the world's most thoroughly vetted form of writing [@cabanac_tortured_2021; @maiberg_ai_2024]. The result is predictable: future GenAI models will inevitably be trained using data that was generated by previous GenAIs, leading to outputs that will be less varied, less diverse, and more norm-conforming. In other words, we may be witnessing _Peak ChatGPT_, and future model versions, instead of improving, may begin to regress to duller versions of their own previous outputs. To compound things, as the backlash against an exploitative data rush is suuported by new tools to generate data that are deleterious to the training process of a GenAIs [@salman_raising_2023], including _PhotoGuard_ [@heikkila_this_2023], _Nightshade_ [@shan_nightshade_2024], and _Shadowcast_ [@xu_shadowcast_2024].

All in all, the evidence suggests that the scarcity of fresh data, paired with more abundant model-generated and/or poisoned data, will make the performance of these machines stagnate or deteriorate.

## What do GenAIs do for the corporations pushing them? {#what-do-ais-for-corporations}

AIs do not offer much of value in teaching and learning: they are wasteful at best and harmful at worst. But the technology must do something for someone, otherwise it would not be ubiquitous. So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities^[Some observers believe that the AI fever has become detrimental to actual innovation [@siegel_ai_2023; @vinsel_dont_2023].]. Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD) [@tricot_venture_2021]. But, the industry has not necessarily made money for investors yet.

@berber_peter_2024, for instance, reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue. @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google. At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 monthly queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) before OpenAI stareted losing money on that one customer. The profitability of AI would look even worse if not for the fact that the cost of labor is contained by outsourcing  to the Global South the most disturbing, degrading, and damaging aspects of training AIs [@nix_analysis_2024].

@@
